---
title: "Introduction to R and the tidyverse"
author: "Paolo Crosetto"
fontsize: 9pt
output:
  ioslides_presentation: default
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
theme_set(theme_minimal())
```

# Lecture 2: data import, export, manipulation, merging, tidying


## Todays topics

> today we will deal with five topics:

1. data import and export: `readr`, `haven`
2. manipulating data: `dplyr`
2. joining data from different tables: `dplyr`
3. *tidying* data: `tidyr`


## Before we start: nycflights

- `install.packages("nycflights13")`
- some data about all flights from New York airports in 2013
- we get to know arrival and departure times, delays, carrier, some info about the plane
- origin, destination, and so on
- not particularly interesting *per se* but big (336K observations)


# importing data

## getting data into R: packages

> up to now we have worked with data sets that come from *packages*

- easy to do: install a package, then call a function with data attached
- all the hard work has been made for you
- if you wish you can import the data into your workspace
- e.g.
```{r}
library(nycflights13)
df <- flights

```

## getting data into R: other sources

> life is not always that easy

- you might have data in the form of (aaarg!) Excel files
- you might have comma separated (csv) data
- you might have data coming from SPSS, SAS, STATA, ...
- or text data from ASCII sources

```{r, echo=FALSE, results='hide'}
rm(df)
```


## getting data into R: `readr` vs `haven`

- when you load the tidyverse (`library(tidyverse)`) you automatically load `readr`
- this is a package that gives you (verb) functions to load data into R nicely
- `readr` provides functions to load most text-based delimited files
- especially .csv
- if you want to read in a STATA or SAS or SPSS file, you need the package `haven` (`library(haven)`)
- `readr` is autmatically loaded by the *tidyverse* call
- `haven` needs to be loaded explicitely (not shown here)

## A simple example

> you find some data here: https://goo.gl/kPycfH

- this is the human develoment index, by country
- highest numbers (nearest to 1) are better

- save the file to disk to somewher you know about
- save it as HDI.csv
- open it up with a text editor: what do you see?

## A simple example

> now that your data is saved, how do you import it to R?

- you use `read_csv("path_to_file")`
- in my case: 

```{r}
df <- read_csv("/home/paolo/Dropbox/Public/HDIdata.csv")
```


## there is more but...

> `read_csv` just made under the hood a ton of things for you

- but it doesn't really matter at your stage
- so you just live with the results. 
- other useful functions:
- if the separator is `;` rather than `,` use `read_csv2`
- if the separator is a TAB rather than `,` use `read_tsv`

## some hints

- you can always export to .csv in all programs
- *even in Excel!*
- so once you have exported to .csv, all is downhill from there
- and it is even ebtter to do it because .csv is universal
- while other binary formats (.dta, .xls...) force you to have the appropriate tool for reading them
- so try to keep a copy of your data in a text-based format, it is always readable should everything go wrong.

# Data manipulation

## Getting to know the data: inspection

```{r}
library(nycflights13)

flights
```

## Getting to know the data: View

```{r}
View(flights)
```

- `View()` opens an Rstudio data window
- in that window you can 
- sort
- arrange
- inspect variables


## Inspecting data and summary statistics

```{r}
summary(flights)
```


## Importing data in your workspace

- `flights` is not yet in your workspace
- *if* you want to import it, you have to do it explicitely
- using `<-`
- the standard name for a data frame in usual code is `df` (you are free to use any other)

```{r}
df <- flights
df
```


## Data manipulation: **dplyr**

> we will use the package **dplyr**, from the tidyverse

- don't worry about the strange name, it has a reason. 
- install it with the usual `install.packages("dplyr")`
- but it should be already there from last (included in `tidyverse`)

## Structure of **dplyr**

- the idea is to have simple, direct *verbs* to do the main jobs for you
- dplyr verbs take a data.frame as a first argument
- do some manipulations on it
- and always return a data.frame
- they do **not** alter the data saved on memory
- which is **good**, because you can manipulate data without altering it
- if you want to save the altered dataset, use assign `<-`

## Data manipulation: **filter()**

> **filter()** allows to extract **rows** from the data frame

- `filter()` takes two arguments:
- `filter(data, logic)`
- `data` is your data 
- `logic` is a  logical statement that tells `filter()` what mut be included or not
- R understands all usual comparison operators `>, >=, <, <=, !=, ==`
- R understand also logial operators are `&, |, !`

## Filtering

> all December flights

```{r}
filter(df, month == 12)
```

## Filtering over multiple criteria
> all Christmas flights departed around midday

```{r}
filter(df, month == 12 & day == 25 & dep_time>1100 & dep_time <1300)
```

## multiple **or** statements and `%in%`

- mutliple 'or' statement can be tricky
- you want all summer flights (June, July, August)
- then you should do `filter(df, month == 6 | month == 7 | month == 8)`
- can  become cumbersome
- you can use an alternative notation, `filter(df, month %in% c(6,7,8))`

## Sort data: `arrange()`
```{r}
filter(df,month == 7 & day == 17 & origin == 'LGA')
```

- `arrange()` lets you sort data
- it is like the sorting you do in `View()`, but:
- 1. it is done on the console, and
- 2. it lets you save the new order to a data frame
- this can be useful when looking for special observations, like the first flight of each day


## Selecting columns: `select()`

> What `filter()` does with rows, `select()` does with columns

- you want more than one variable: you list all variables by bare name: `select(data, var1, var2, var3, ...)`
- you want to exclude (drop) some variables: you use the minus sign: `select(data, -var1, -var2, ...)`
- you want to exploit naming patterns: you use `starts_with("string")`, `ends_with("string")`, `contains("string")`
- you want to select all variables: you use `everything()` 
- `everything` is useful to *reorder variables*


## renaming variables: `rename()`

- `rename()` is a version of `select()` that keeps all variables
- you use it as `rename(data, newname = oldname)`

```{r}
rename(df, mois = month)
```


## creating new variables: `mutate()`

> you want to create a *new* variable with some manipulation

- you can use `mutate(data, newvar = f(oldvar))`
- where `f()` is some function or manipulation

```{r}
df_new <- mutate(df, starting_year = year - 2013)
select(df_new, starting_year)
```


## `mutate()` properties

- you can create more variables in one single `mutate()` call

```{r}
df_new <- mutate(df, starting_year = year - 2013, speed = distance /air_time * 60,
                 speed2 = speed**2)
select(df_new, starting_year, speed, speed2)
```


## `mutate()` properties

- you can *use* right away the newly created variables

```{r}
df_new <- mutate(df, speed = distance /air_time * 60, speed_squared = speed ** 2)
select(df_new, speed, speed_squared)

df %>% 
  filter(day == 25 & month == 12) %>% 
  group_by(origin) %>% 
  summarize(meandelay = mean(dep_delay, na.rm = T))

df %>% 
  group_by(month, carrier) %>% 
  summarise(meandelay = mean(dep_delay, na.rm = T))

```


## functions that work with `mutate()`

- all *vector* functions: input a vector, output a vector
- arithmetic operations: + - * / ^ (or **) exp(), sqrt(), log()
- offset: lag() to refer to the -1 period. *useful for increments*
- cumulative funcsions: `cumsum()`, etc...
- logical (see `filter()`)

## summarise your data: `summarize()`

- if you want to create aggregations and summary fo your data
- you cna use `summarize(data, newvar = f(oldvar))`
- it works similarly to `mutate()`  but it returns a **scalar**

```{r}
summarize(df, meandelay = mean(dep_delay, na.rm = TRUE))
```

## `summarize()` properties

- you can do more than one summary in one go

```{r}
summarize(df, meandelay = mean(dep_delay, na.rm = TRUE), sddelay = sd(dep_delay, na.rm = TRUE), meandeptime = mean(dep_time, na.rm = TRUE))
```

## making `summarize()` more useful: `group_by()`

> `summarize()` is not so useful -> one value

- but it can be extremely useful if we can group data by a special variable, to see, e.g., monthly values
- this is what you do with `group_by()`
- which groups the data frame by levels of one (or more) variablesn
- nothing changes but now R knows this is a grouped data frame

```{r}
group_by(df, month)
```


## example: delay by month

- output has 12 rows

```{r}
df_bymonth <- group_by(df, month)
summarize(df_bymonth, meandelay = mean(dep_delay, na.rm = TRUE))
```

## example: delay by month **and** day

- output has 365 rows

```{r}
df_bymonth_byday <- group_by(df, month, day)
summarise(df_bymonth_byday, meandelay = mean(dep_delay, na.rm = TRUE))
```


## Exercise: a complex example

> **what is the speed of flights departing around midday (11 - 13), by month?**

- the input is the flights df
- use `filter()`, `select()`, `mutate()`, `group_by()` and `summarize()`
- the final output is a df with 2 variables (month and speed) and 12 observations (one for each month)

> *à vous de jouer!*


## A somewhat more complex example

- let's combine `filter()`, `select()`, `mutate()`, etc...

- we want to know the average speed of flights departing around midday, by month

```{r, results=FALSE}
df_midday <- filter(df, dep_time > 1100 & dep_time < 1300)
df_midday_reduced <- select(df_midday, month, air_time, distance)
df_midday_speed <- mutate(df_midday_reduced, 
                          speed = distance / air_time * 60)
df_midday_grouped <- group_by(df_midday_speed, month)
df_midday_final <- summarise(df_midday_grouped, 
                             meanspeed = mean(speed, na.rm = TRUE))
```

```{r}
df_midday_final
```

## Multiple operations: problems

> the code in the previous slide has several problems

- it forces you to create several different data frames
- this clutters your environment
- but not only: at each step, you have to refer to the *right* data frame
- else you make mistake
- hard to find proper names
- easy to make errors
- what can we do?

## Enters the pipe! `%>%`

> the *pipe* operator

- feeds the *result* of a command as the *first argument* of the following command
```{r}
2*2
exp(4)
(2*2) %>% exp()
```
- since *dplyr* verbs always return a data frame, and always have a data frame as first input, this saves your life

## the same code with the pipe
```{r}
df_midday  %>%  filter(dep_time > 1100 & dep_time < 1300) %>% 
                select(month, air_time, distance) %>% 
                mutate(speed = distance / air_time * 60) %>% 
                group_by(month) %>% 
                summarise(meanspeed = mean(speed, na.rm = TRUE))
```

## Learning to use the pipe

- what is the mean delay of flights by carrier, for each month?
- what is the maximum departure delay occurred for each of the three NYC airports, by each day?
- what is the mean air time for each of these three airports? from which airport do the longer haul flights depart?