---
title: "Exam"
author: "Morghi Younesse"
date: "09/11/2021"
output:
  slidy_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.


## Slide with R Output
 
```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(rvest)

load("C:/Users/you_n/Desktop/M1/R 2022/Intro-to-R-M1-C2ES-2021/my_data_frame.Rda")

SW <- swim_years
#garder les colones importantes 
SW = SW %>%
  select(-type,-country,-grp_id)
#suppression des NA 
SW <- SW %>% 
  drop_na()


```




## Slide with Plot

```{r }
plot(pressure)
```



```{r}

```

